---
title: "SBM output analysis: CPTAC (BRCA) data"
author: "Anne-Claire Kroger, anne-claire.kroger@curie.fr"
date: "2024-06-28"
output: html_document
---

# Read the SBM output and do some first visualization of the results

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(reshape2)
library(ggplot2)
library(matrixStats)
library(pheatmap)
library(dplyr)
library(networkD3)
library(stringr)
library(fossil)

source("sbm_output_analysis_functions.R")
```

### Import published clustering information

```{r}
brca_cluster <- read.csv("data/input_brca_cptac/BRCA_meta_paper_clustering.csv")
rownames(brca_cluster) <- brca_cluster$Sample.ID

```

## Single-omics hSBM

### RNA 4k features

Loading all the results from hSBM using the custom function. Results
will be saved in the right structure in the specifiied folder names if
using the python script (see script 02). 

* folder_path - Path to the
folder with saved results as specifified in the python script 02 
* level - level of the SBM output to be loaded 
* type - name used to save
the "worddist" matrices in python, as the SBM algorithm doesn't do it in
the save.data() function like the rest of the output 
* omics_leyers - number of omics layers used in this analysis, default is 1

```{r}
#load data
brca_rna_4k_l0 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_rna_tests_4k", level = 0, type = "brca_rna_4k")
brca_rna_4k_l1 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_rna_tests_4k", 1, type = "brca_rna_4k")
brca_rna_4k_l2 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_rna_tests_4k", 2, type = "brca_rna_4k")

```

The output object contains all relevant outoput for one level: 

* clusters - The discrete assignment of samples to a cluster. On each
increasing level of the calculated SBM model, the number of clusters
decreases, level 0 has the lowest number of clusters, often matching the
original number of samples 
* topics - The discrete assignment of
features to topics, also decreases in higher SBM levels. In a
multi-omics nSBM model, there are different topics for each omics type.
Careful, the name of the topics is just a simple numbering, meaning they
are the same for each omics layer (make sure to add an additional
identifier when working with topics of different layers together). 
* todist - The probablity distribution for a sample to be characterized by
a certain topic (different for each omics layer). This can be seen as a
reduced dimension, with topics as the new lower dimension features 
* worddist: The probability distribution for a feature to be part of a
topic (different for each omics layer). Is the value for a feature
higher the 0, it means that this feature is assigned to a specific
topic. A feature only has a value for one topic, since they are
exclusive.

```{r}
as.data.frame(brca_rna_4k_l0$summary)
as.data.frame(brca_rna_4k_l1$summary)
as.data.frame(brca_rna_4k_l2$summary)
```

Check out the cluster numbers through the summary list in each sbm
object. Level 1 has 4 clusters which matches the expected number of
clusters for the cancer type.

We can visualize the results through a sankey plot for several levels or
a barplot for one level. In both cases we directly compare the results
to the previously published annotations

```{r}
#sankey plot l1 & l2

brca_samp <- as.list(brca_cluster$Sample.ID)

#using level 1 & 2 in this case
l1 <- unlist(lapply(brca_samp, function(samp) {which(brca_rna_4k_l1$clust == samp, arr.ind = TRUE)[2]}))
l2 <- unlist(lapply(brca_samp, function(samp) {which(brca_rna_4k_l2$clust == samp, arr.ind = TRUE)[2]}))

#add the prviously published annotation "NMF.Cluster"
df <- data.frame("level_2" = l2, "level_1" = l1, "brca_subgroup" = brca_cluster$NMF.Cluster, row.names = brca_cluster$Sample.ID)
df_s <- get_sankey_input(df)
p <- sankeyNetwork(Links = df_s$links, Nodes = df_s$nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE)


p

```

```{r}
#barplot level 1

df <- get_nsbmclust_df(brca_rna_4k_l1$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)

df$cluster <- rownames(df)
df <- melt(df, id = "cluster")

ggplot(df, aes(fill=variable, y=value, x=cluster)) + 
    geom_bar(position="stack", stat="identity")

```

### Prot 2k feature

```{r}
#load data
brca_prot_2k_l0 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_prot_tests_2k", 0, type = "brca_prot_2k")
brca_prot_2k_l1 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_prot_tests_2k", 1, type = "brca_prot_2k")
brca_prot_2k_l2 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_prot_tests_2k", 2, type = "brca_prot_2k")

```

```{r}
as.data.frame(brca_prot_2k_l0$summary)
as.data.frame(brca_prot_2k_l1$summary)
as.data.frame(brca_prot_2k_l2$summary)
```

```{r}
#sankey plot l1 & l2

brca_samp <- as.list(brca_cluster$Sample.ID)

#using level 1 & 2 in this case
l1 <- unlist(lapply(brca_samp, function(samp) {which(brca_prot_2k_l1$clust == samp, arr.ind = TRUE)[2]}))
l2 <- unlist(lapply(brca_samp, function(samp) {which(brca_prot_2k_l2$clust == samp, arr.ind = TRUE)[2]}))

#add the prviously published annotation "NMF.Cluster"
df <- data.frame("level_2" = l2, "level_1" = l1, "brca_subgroup" = brca_cluster$NMF.Cluster, row.names = brca_cluster$Sample.ID)
df_s <- get_sankey_input(df)
p <- sankeyNetwork(Links = df_s$links, Nodes = df_s$nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE)


p

```

```{r}
#barplot level 1

df <- get_nsbmclust_df(brca_prot_2k_l1$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)

df$cluster <- rownames(df)
df <- melt(df, id = "cluster")

ggplot(df, aes(fill=variable, y=value, x=cluster)) + 
    geom_bar(position="stack", stat="identity")

```

### Phospho 8k features

```{r}
#load data
brca_phos_8k_l0 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_phos_tests_8k", 0, type = "brca_phos_8k")
brca_phos_8k_l1 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_phos_tests_8k", 1, type = "brca_phos_8k")
brca_phos_8k_l2 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_phos_tests_8k", 2, type = "brca_phos_8k")
brca_phos_8k_l3 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_phos_tests_8k", 3, type = "brca_phos_8k")

```

```{r}
as.data.frame(brca_phos_8k_l0$summary)
as.data.frame(brca_phos_8k_l1$summary)
as.data.frame(brca_phos_8k_l2$summary)
as.data.frame(brca_phos_8k_l3$summary)
```

```{r}
#sankey plot l1 & l2

brca_samp <- as.list(brca_cluster$Sample.ID)

#using level 1 & 2 in this case
l1 <- unlist(lapply(brca_samp, function(samp) {which(brca_phos_8k_l1$clust == samp, arr.ind = TRUE)[2]}))
l2 <- unlist(lapply(brca_samp, function(samp) {which(brca_phos_8k_l2$clust == samp, arr.ind = TRUE)[2]}))
l3 <- unlist(lapply(brca_samp, function(samp) {which(brca_phos_8k_l3$clust == samp, arr.ind = TRUE)[2]}))


#add the prviously published annotation "NMF.Cluster"
df <- data.frame("level_3" = l3, "level_2" = l2, "level_1" = l1, "brca_subgroup" = brca_cluster$NMF.Cluster, row.names = brca_cluster$Sample.ID)
df_s <- get_sankey_input(df)
p <- sankeyNetwork(Links = df_s$links, Nodes = df_s$nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE)


p

```

```{r}
#barplot level 1

df <- get_nsbmclust_df(brca_phos_8k_l1$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)

df$cluster <- rownames(df)
df <- melt(df, id = "cluster")

ggplot(df, aes(fill=variable, y=value, x=cluster)) + 
    geom_bar(position="stack", stat="identity")


```

### Further analysis for any omics analysis level of interest

(can also be copy and pasted under each analysis)

```{r}
#define omics type and matching level of interest
loi <- brca_rna_4k_l1

loi_name <- "hSBM l=1 brca rna 4k"
```

Calculate the rand index

```{r}
df <- get_nsbmclust_df(loi$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)
rand_index_sbmcluster(df)
```

Take a closer look at topics by using the analyse_topics function which
calculates different normalisations for topic scores

```{r}
stats_loi <- analyse_topics(loi)

```

Check the centered topic values aka "importance" for each each topic
using the SBM assigned clustering. Centered topic values per cluster are
calculated according to this publication:
<https://www.mdpi.com/2072-6694/16/7/1350>

```{r}
ggplot(stats_loi$info_centered, aes(x=cluster, y=value))+
  geom_boxplot() +
  geom_point(aes(group = factor(cluster), color=topic_length), show.legend = T, size = 2) +
  labs(x="", y="Centered topic importance", title = paste("Results", loi_name)) +
  theme(plot.title = element_text(hjust = 0.5))

```

Check correlation between centered topic importance and topic length, vs
correlation when only using mean topic importance without centering by
cluster.

```{r}
cluster <- unique(stats_loi$info_centered$cluster)


for (cl in cluster) {
  p <- ggplot(stats_loi$info_centered[stats_loi$info_centered$cluster==cl,], aes(x=value, y=topic_length)) +
    geom_point() +
    labs(x="Centered topic importance", y="Topic size", title = paste("Results",loi_name, cl)) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
  print(p)
} 



p <- ggplot(stats_loi$info_centered, aes(x=topic_mean_imp, y=topic_length)) +
    geom_point() +
    labs(x="Mean topic importance", y="Topic size", title = paste("Results", loi_name)) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
print(p)


```

Correlation value can also be simply calculated without visualization.

```{r}
topic_length_corr(stats_loi)
```

## Multi-omics nSBM

### Medium size data, all 3 layers (4k RNA, 2k Prot, 8k Phos)

Load multi-omics results, but specify the number of omics. The output
has the same structure but with additional matrices for topics,
topicdist and worddist for every omics layer. The clustering is the
integratd result with only one clustering using all omics layers.

```{r}
#brca_medium_l0 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium", 
#                                 level = 0, 
#                                 omics_layers = 3)

brca_medium_l1 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium", 
                                 level = 1, 
                                 omics_layers = 3)

brca_medium_l2 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium", 
                                 level = 2, 
                                 omics_layers = 3)

```

```{r}
t(as.data.frame(brca_medium_l1$summary))
t(as.data.frame(brca_medium_l2$summary))
```

```{r}
#sankey plot l1 & l2

brca_samp <- as.list(brca_cluster$Sample.ID)

#using level 1 & 2 in this case
l1 <- unlist(lapply(brca_samp, function(samp) {which(brca_medium_l1$clust == samp, arr.ind = TRUE)[2]}))
l2 <- unlist(lapply(brca_samp, function(samp) {which(brca_medium_l2$clust == samp, arr.ind = TRUE)[2]}))


#add the prviously published annotation "NMF.Cluster"
df <- data.frame("level_2" = l2, "level_1" = l1, "brca_subgroup" = brca_cluster$NMF.Cluster, row.names = brca_cluster$Sample.ID)
df_s <- get_sankey_input(df)
p <- sankeyNetwork(Links = df_s$links, Nodes = df_s$nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE)


p

```

```{r}
#barplot level 1

df <- get_nsbmclust_df(brca_medium_l1$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)

df$cluster <- rownames(df)
df <- melt(df, id = "cluster")

ggplot(df, aes(fill=variable, y=value, x=cluster)) + 
    geom_bar(position="stack", stat="identity")


```

### Medium size data without phosph0 (4k RNA, 2k Prot)

```{r}
#brca_medium_noPhos_l0 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium_noPhos", 
#                                 level = 0, 
#                                 omics_layers = 2)

brca_medium_noPhos_l1 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium_noPhos", 
                                 level = 1, 
                                 omics_layers = 2)

brca_medium_noPhos_l2 <- read_sbm_output(folder_path = "data/sbm_output_brca_cptac/brca_nsbm_tests_medium_noPhos", 
                                 level = 2, 
                                 omics_layers = 2)

```

```{r}
t(as.data.frame(brca_medium_noPhos_l1$summary))
t(as.data.frame(brca_medium_noPhos_l2$summary))
```

```{r}
#sankey plot l1 & l2

brca_samp <- as.list(brca_cluster$Sample.ID)

#using level 1 & 2 in this case
l1 <- unlist(lapply(brca_samp, function(samp) {which(brca_medium_noPhos_l1$clust == samp, arr.ind = TRUE)[2]}))
l2 <- unlist(lapply(brca_samp, function(samp) {which(brca_medium_noPhos_l2$clust == samp, arr.ind = TRUE)[2]}))


#add the prviously published annotation "NMF.Cluster"
df <- data.frame("level_2" = l2, "level_1" = l1, "brca_subgroup" = brca_cluster$NMF.Cluster, row.names = brca_cluster$Sample.ID)
df_s <- get_sankey_input(df)
p <- sankeyNetwork(Links = df_s$links, Nodes = df_s$nodes,
                     Source = "IDsource", Target = "IDtarget",
                     Value = "value", NodeID = "name", 
                     sinksRight=FALSE)


p

```

```{r}
#barplot level 1

df <- get_nsbmclust_df(brca_medium_noPhos_l1$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)

df$cluster <- rownames(df)
df <- melt(df, id = "cluster")

ggplot(df, aes(fill=variable, y=value, x=cluster)) + 
    geom_bar(position="stack", stat="identity")


```

### Further analysis for any multi-omics integration analysis level of interest

```{r}
#define omics type and matching level of interest
loi <- brca_medium_noPhos_l1
number_omics <- 2

loi_name <- "nSBM l=1 brca medium no phospho"
```

Calculate the rand index:

```{r}
df <- get_nsbmclust_df(loi$clust, anno = brca_cluster, anno_samplecol = 1, anno_groupcol = 8)
rand_index_sbmcluster(df)
```

Take a closer look at topics by using the analyse_topics function which
calculates different normalisations for topic scores, specifiying the
number of omics layers.

```{r}
stats_loi <- analyse_topics(loi, omics_layers = number_omics)

```

Check the centered topic values aka "importance" for each each topic
using the SBM assigned clustering. Since they are different topics for
each omics layer, they are different topics for each layer. Centered
topic values per cluster are calculated according to this publication:
<https://www.mdpi.com/2072-6694/16/7/1350>.

```{r}
ggplot(stats_loi$rna$info_centered, aes(x=cluster, y=value))+
  geom_boxplot() +
  geom_point(aes(group = factor(cluster), color=topic_length), show.legend = T, size = 2) +
  labs(x="", y="Centered topic importance", title = paste("Results", loi_name, ", RNA topics")) +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(stats_loi$prot$info_centered, aes(x=cluster, y=value))+
  geom_boxplot() +
  geom_point(aes(group = factor(cluster), color=topic_length), show.legend = T, size = 2) +
  labs(x="", y="Centered topic importance", title = paste("Results", loi_name, ", Proteomics topics")) +
  theme(plot.title = element_text(hjust = 0.5))

if (number_omics == 3) {
  ggplot(stats_loi$phos$info_centered, aes(x=cluster, y=value))+
  geom_boxplot() +
  geom_point(aes(group = factor(cluster), color=topic_length), show.legend = T, size = 2) +
  labs(x="", y="Centered topic importance", title = paste("Results", loi_name, ", Phospho topics")) +
  theme(plot.title = element_text(hjust = 0.5))
}

```

Check correlation between centered topic importance and topic length, vs
correlation when only mean topic importance without centering.

```{r}
cluster <- unique(stats_loi$rna$info_centered$cluster)


for (cl in cluster) {
  p <- ggplot(stats_loi$rna$info_centered[stats_loi$rna$info_centered$cluster==cl,], aes(x=value, y=topic_length)) +
    geom_point() +
    labs(x="Centered topic importance", y="Topic size", title = paste("Results",loi_name, cl, "RNA Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
    print(p)
  
  l <- ggplot(stats_loi$prot$info_centered[stats_loi$prot$info_centered$cluster==cl,], aes(x=value, y=topic_length)) +
    geom_point() +
    labs(x="Centered topic importance", y="Topic size", title = paste("Results",loi_name, cl, "Proteomics Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
  print(l)
  
  
  if (number_omics == 3) {
   o <- ggplot(stats_loi$phos$info_centered[stats_loi$phos$info_centered$cluster==cl,], aes(x=value, y=topic_length)) +
    geom_point() +
    labs(x="Centered topic importance", y="Topic size", title = paste("Results",loi_name, cl, "Phospho Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
  }
} 



ggplot(stats_loi$rna$info_centered, aes(x=topic_mean_imp, y=topic_length)) +
    geom_point() +
    labs(x="Mean topic importance", y="Topic size", title = paste("Results", loi_name, ", RNA Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))

ggplot(stats_loi$prot$info_centered, aes(x=topic_mean_imp, y=topic_length)) +
    geom_point() +
    labs(x="Mean topic importance", y="Topic size", title = paste("Results", loi_name, ", Proteomics Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))

if (number_omics == 3) {
  ggplot(stats_loi$phos$info_centered, aes(x=topic_mean_imp, y=topic_length)) +
    geom_point() +
    labs(x="Mean topic importance", y="Topic size", title = paste("Results", loi_name, ", Phospho Topics")) +
    geom_smooth(method='lm', formula= y~x) +
    theme(plot.title = element_text(hjust = 0.5))
}


```

Correlation value can also be simply calculated without visualization.

```{r}
topic_length_corr(stats_loi$rna)
topic_length_corr(stats_loi$prot)
# topic_length_corr(stats_loi$phos)
```

## Identify topics of interest

This is a a simple way to identify topics of interest as specified in
this publication: <https://www.mdpi.com/2072-6694/16/7/1350>.

```{r}
#for rna
stats <- stats_loi$rna$info_all[,10:12]
pheatmap(stats, main = paste(loi_name, ", RNA Topics"))

m <- apply(stats, 2, mean)
std <- apply(stats, 2, function(x){sqrt(var(x))})
cl <- stats > m+2*std
cl_rna <- rownames(cl)[rowSums(cl) != 0]

pheatmap(stats[cl_rna,], main =  paste(loi_name, ", RNA Significant Topics"))

#for proteomics
stats <- stats_loi$prot$info_all[,10:12]
pheatmap(stats, main = paste(loi_name, ", Proteomics Topics"))

m <- apply(stats, 2, mean)
std <- apply(stats, 2, function(x){sqrt(var(x))})
cl <- stats > m+2*std
cl_prot <- rownames(cl)[rowSums(cl) != 0]

pheatmap(stats[cl_prot,], main =  paste(loi_name, ", Proteomics Significant Topics") )



```

Topics for each cluster can also be identified manually, depending on
the results and number of overall Topics/ Clusters.

```{r}
todist_cut_rna <- brca_medium_noPhos_l1$todist_rna[,cl_rna]
colnames(todist_cut_rna) <- paste0(colnames(todist_cut_rna), "_T") #adding T for transcriptomics and P for proteomics to distinguisch the different topics
pheatmap(t(todist_cut_rna), annotation = brca_cluster[,c(7,8)])

todist_cut_prot <- brca_medium_noPhos_l1$todist_prot[,cl_prot]
colnames(todist_cut_prot) <- paste0(colnames(todist_cut_prot), "_P")
pheatmap(t(todist_cut_prot), annotation = brca_cluster[,c(7,8)])

todist_c <- cbind(todist_cut_prot, todist_cut_rna)
colnames(todist_c) <- paste0(str_remove(colnames(todist_c), "_.*$"), str_extract(colnames(todist_c), "_."))

pheatmap(t(todist_c), annotation = brca_cluster[,c(7,8)], treeheight_row = 0, show_colnames = FALSE)
```

```{r}
sessionInfo()
```
